Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models

### 2018-01-07 paper-amia-summit-2018/amia_summit_2018-submitted.tex
fixed typo and expand PPC term

\title{Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences}

@@

\institutes{
$^1$Department of Computer Science,  
$^2$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics. In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges. Our method achieved 70.38\%, 60.67\%, 86.77\% F1-score for under-sampled sequences and 77.75\%, 75.20\%, 83.81\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively. The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. Sequence classification methods are utilized for classifying temporally ordered sequences. Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}. In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease. Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.

%Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection. In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}. Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. 
%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. 
Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.  
%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information is learned from the dataset and used as a ``model'' to classify sequences.

%This makes sequence classification a more challenging task than traditional classification.

One of the model-based classification methods is a probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM). Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step. Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory. It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}. A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}. In Lipton et. al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc. LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit. A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed. These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.

Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

%This work is the next step in a series of experiments to develop the [informatics based methods described above]. 
In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
We propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into client codes (adolescent and caregiver) group and counselor codes group. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Here, the successful sequences had an average length of 9.79 while that of unsuccessful sequences was 9.65. For each of the probabilistic models (Markov chain and HMM), two models were trained, one model was trained using successful sequences and another one was trained using unsuccessful sequences.

@@

Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level. In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity dataset. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.

A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end. It was noticed that same context ``Yeah" and ``Yes'' represent different behavior codes CHT+ and HUPW, respectively. Because human brain can remember the past behaviors, and classify the current context into different behavior codes after considering their past behaviors. For this reason, we only consider those models that can capture the past events for the classification of behavior sequences.\\

@@

\subsection*{\textit{Prediction method}}
Generally, a sequence is a temporally ordered set of events. In this study, an event is a behavior code that also has a symbolic representation, such as $LUP+$ (low uptake, positive), $OQECHT+$ (open-ended question, elicit change talk positive), etc.  Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some part of a motivational interview, the task of predicting interview success can be viewed as sequence classification. Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are ``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, won2004training}. However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variants of RNN models in our experiments: long short-term memory and gated recurrent unit.

\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state but not on its past states (Markov property). For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, LUP+, LUP-, HUPW, OQO, CQTBN, CQECHT+,...\}$ is a set of vertices, consisting of adolescent, caregiver and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
\begin{equation}
p(S\rightarrow successful) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow successful) > 0 $, the interview is predicted to result in positive change talk or commitment language. Otherwise, it would be classified as a negative change talk or commitment language.

The above model also referred as first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities. In our experiment, we also considered second-order Markov model that compute transition probabilities based on previous two states.

@@

\textbf {Embedding}: We took the inspiration for the representation of behavior codes from the idea of word embedding \cite{bengio2003neural}. Word embedding is a representation of words in low-dimensional space by vectors which contain the feature of the words. In our study, we employed embedding in place of one-hot vectors for the representation of behavior code since one-hot vectors are high-dimensional and sparse. Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations. The embedding was learned during the training of the classification model. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embedding in 44-dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster at the left of the figure. The nearest neighbors of CQECML+ are highlighted by their color intensity i.e. OQECML+ is more purple means closer to CQECML+. The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.

@@

\section*{Results}
Experimental evaluation of the proposed method is conducted on both under and over-sampled sequences. Predictive performance summary of the proposed methods on under and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\ 
\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of under and over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Undersampling}} & \multicolumn{3}{|c|}{\textbf{Oversampling}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

>>>>

\title{Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models}

@@

\institutes{
$^1$Department of Computer Science, Wayne State University, Detroit, Michigan \\  
$^2$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
}

@@

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences generated by a physiological, genomic or psychological process to make predictions about the outcome of that process arises in
many domains of clinical informatics. In this paper, we focus on the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviewing.
Specifically, we formulate the problem of predicting the success of a motivational interview (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of
coded patient-provider communication exchanges as binary sequence classification.
%Severe imbalance between positive and negative training sequences is an additional challenge that arises in the context of motivational interviewing.
We propose two methods that are based on deep learning, such as Recurrent Neural Networks (RNNs), and probabilistic models, such as Markov Chain (MC) and Hidden Markov Model (HMM), and compare the
accuracy of these methods using communication sequences annotated with behavior codes from real-life motivational interviews. Our experiments indicate that the deep learning-based method is
significantly more accurate than the methods based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM,
respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate
that the proposed method can be used for real-time monitoring of the progression of clinical interviews and identification of the most effective communication strategies, which significantly decreases
the effort required to develop psychological interventions and increases their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by genomic, psychological or psychological processes arise in many different domains of clinical informatics (e.g.,
DNA sequences, gene activations, ECG measurements, laboratory results, diagnostic codes, utterances during clinical interviews). Classification (or categorization) is a type of analysis of those
sequences that has a broad range of important practical applications, from protein classification \cite{yakhnenko2005discriminatively} and function prediction \cite{deshpande2002evaluation} to
detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as their temporal order, makes sequence classification a
more challenging task than traditional classification. Predicting the outcome of those sequences (e.g. physiological or behavior response) can also be formulated as sequence classification.

In general, sequence classification methods fall into one of the three major categories: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature
vector and apply a standard supervised learning methods, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
Feature-based methods for sequence classification have had limited success, since traditional bag-of-words feature representation methods cannot easily account for the order of observations in a
sequence.
%Shapelet \cite{ye2009time} and pattern based techniques \cite{kudenko1998feature, lesh1999mining} as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of
%standard classifiers as well.
Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is
Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by time points.
The third type of sequence classification methods first creates a probabilistic model, such as Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class
using the training data and then classifies new sequences by applying the constructed models. MC and HMM explicitly capture short-term temporal dependencies (e.g. first- and second-order MC and HMM
\cite{kundu1988recognition}) between adjacent observations in a sequence. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short-
and long-term dependencies and were shown to be effective for modeling different types of sequential data. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which
successfully addresses the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and
handwriting \cite{nion2013handwritten} recognition to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to
recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting
the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
Unit\cite{chung2014empirical} (GRU), was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.

In this paper, we address the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of
Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that
approximately one-third (31.8\%) of US children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and
have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI),
an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires,
ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior
change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of
patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since
manual examination and sequential pattern analysis of transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years.
Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the
mechanisms of MI's efficacy, in particular.

Our goal is to test the applicability of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for predicting the success of clinical interviews (i.e. eliciting a
particular type of patient behavioral response, such as CHT) and dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on a sequence of
coded previous patient-provider communication exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored
several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook
\cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability
of state-of-the-art methods for sequence modeling to analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavior in a context of motivational
interviews, in particular.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists
of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, in which the codes are grouped into
patient (adolescent and caregiver) codes and counselor codes. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in
positive change talk and commitment language statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the
sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Successful sequences had an average length of
9.79 utterances, while unsuccessful sequences had on average 9.65 utterances. For each of the probabilistic models (MC and HMM), two models were trained, one model was trained using successful
sequences and another one was trained using unsuccessful sequences.

@@

Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the
performance of an employed algorithm for an imbalanced data, since simply predicting the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to
handle class imbalance properly. Many solutions have been proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level. In this study, we
utilized under-sampling and over-sampling for balancing the samples in different classes in our dataset. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used
method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between
minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the
clusters of samples by the cluster centroid in a $k$-means clustering algorithm.

A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Annotation column shows successful sequences of behavior codes from top to bottom. For example,
($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end. As
can be seen, similar utterances, such as  ``Yeah" and ``Yes'', can have different behavior codes (CHT+ and HUPW, respectively) depending on the context.\\

@@

\subsection*{\textit{Prediction method}}
Generally, a sequence can be viewed as a temporally ordered set of events. In this study, an event is a behavior code that also has a symbolic representation, such as $LUP+$ (low uptake, positive),
$OQECHT+$ (open-ended question, elicit change talk positive), etc.  Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some
part of a motivational interview, the task of predicting interview success can be viewed as sequence classification. Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are
``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first and second-order Markov Chain and
Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classify each test sequence based on the maximum
likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence
modeling \cite{mutsam2016maximum, won2004training}. However, the latest advances in deep learning indicate that RNNs provide better results than conventional machine learning methods for the task of
sequence classification. Therefore, we employed two special variants of RNN in our experiments: Long Short-Term Memory (LSTM) \cite{hochreiter1997long} and Gated Recurrent Unit (GRU)
\cite{chung2014empirical}.

\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state but not on its past states (Markov property). For the sequential analysis,
we built two Markov models $M$ and $\overline{M}$, summarizing counselor strategies and patient responses, in the cases of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A
Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, LUP+, LUP-, HUPW, OQO, CQTBN, CQECHT+,...\}$ is a set of vertices, consisting of adolescent, caregiver and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point as:
\begin{equation}
p(S\rightarrow successful) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow successful) > 0 $, a communication sequence is predicted to result in positive change talk or commitment language. Otherwise, it would be classified as a negative change talk or
commitment language.

The above model also referred as first-order MC, since it only considers immediately preceding behavior code, when computing the state transition probabilities. In our experiment, we also considered
second-order Markov model that computes transition probabilities based on previous two states.

@@

\textbf {Embeddings}: We took the inspiration for the representation of behavior codes from the idea of word embeddings\cite{bengio2003neural}. Word embedding is a representation of words in
low-dimensional space by vectors, which contain the features of the words. In our study, we employed embedding in place of one-hot vectors for representation of behavior codes as input to LSTM and
GRU, since one-hot vectors are high-dimensional and sparse. Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by
retaining their internal relations. The embedding was learned during the training of the classification model. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embedding in 44-dimensional
vector space visualized by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left
part of the figure. The nearest neighbors of CQECML+ are highlighted by their color intensity (i.e. OQECML+ being more purple means closer to CQECML+). The right part of the figure demonstrates
another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended
to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.

@@

\section*{Results}
Experimental evaluation of the proposed method is conducted on both under and over-sampled sequences. Predictive performance summary of the proposed methods on under and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\ 
\begin{table}[ht]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of under and over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Undersampling}} & \multicolumn{3}{|c|}{\textbf{Oversampling}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7060 & 0.7044 & 0.7038 & 0.7932 & 0.7799 & 0.7775 \\ \hline
 Markov Chain $2^{nd}$ Order & 0.6395 & 0.6385 & 0.6379 & 0.7111 & 0.7029 & 0.7000\\ \hline
 Hidden Markov Model & 0.6244 & 0.6143 & 0.6067 & 0.7775 & 0.7567 & 0.7520\\ \hline
 LSTM RNN & 0.8672 & 0.8626 & 0.8622 & 0.8411 & 0.8372 & 0.8368\\ \hline
 LSTM RNN - TR & \textbf{0.8733} & \textbf{0.8681} & \textbf{0.8677} & \textbf{0.8424} & \textbf{0.8385} & \textbf{0.8381}\\ \hline
 GRU RNN & 0.8674 & 0.8648 & 0.8646 & 0.8379 & 0.8342 & 0.8337\\ \hline
 GRU RNN - TR & 0.8705 & 0.8676 & 0.8673 & 0.8412 & 0.8377 & 0.8373\\ \hline 
  \end{tabular}
\end{table}  
\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used. Five major conclusions can be drawn
from the results in Table~\ref{tab:result_under_over_sampled}. First, recurrent neural networks outperform probabilistic models and achieve 16.39\%-26.1\% higher F1-score. Second, LSTM with
target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681. Third, target replication strategy improves the
performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics
for under-sampled sequences. Fifth, second-order MC has lower precision, recall, and F-measure than first-order MC. In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 
9.36\%, when going from first to second-order MC model.

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset. For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size
55. Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%,
2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain. Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with
target replication has the highest F1-score among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences, while the performance of
probabilistic models increases. We also observed that all models have the largest value in precision compared to other performance metrics.
